\sys: Controllable Program Regeneration

### 2025-06-14 paper/draft.tex
Add cross-language citations
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>
Improvements in intro
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{abstract}
\end{abstract}

% Given a source program and a declarative user query $f(P, P')$,
% \sys orchestrates a property-guided regeneration process.
% Adapters extract relevant properties from the original program
% using pluggable analyzers,
% which may be language-agnostic (\eg I/O tracing)
% or language-specific (\eg AST parsing),
% and may require execution environments (\eg sandboxing).
% The logic engine formulates a minimum-satisfiability (MinSAT) problem
% over the property space using a plugin-extensible knowledge base that encodes
% compatibility constraints (\eg which transformations preserve purity, which
% properties can be reused across languages).
% The synthesis engine uses these extracted properties to generate candidate
% programs, which are verified by verifier plugins.
% The result is a program $P'$ that satisfies the user's transformation goal
% while preserving required behaviors of the original program.

\maketitle

@@

% TODO: Introduce LLMs around here
Recent advances in large language models (LLMs) have opened new possibilities for automating code transformation tasks.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
This flexibility has positioned them as attractive tools for software evolution.
However, their outputs are shaped by surface-level patterns in the input~\cite{yang2025evaluatinggeneralizationcapabilitieslarge}.
They struggle with negative constraints (\eg \say{do not use this interface}), often overfit to spurious input cues, and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode,jiang2024llmsdreamelephantswhen,hwang2024thinkpinkelephant,roh2025breakthechainreasoningfailuresllms}.
As a result, while LLMs can accelerate prototyping, developers remain unable to reliably steer or verify their correctness.

@@

Despite their power, LLMs and associated tooling remain fundamentally
procedural and ad hoc: developers must prompt and steer them indirectly, often
relying on brittle heuristics or manual trial-and-error to align outputs with
intent.
This process is brittle, hard to audit, and difficult to integrate into systematic workflows.
Bridging this gap calls for a declarative interface for transformation, one that allows explicit semantic constraints and transformation goals, and supports reasoning about their satisfaction.

@@

\heading{Deployment scenarios and limitations}
\sys can be deployed at various points in the software lifecycle.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating third-party modules that must be adapted to local conventions or hardened against security vulnerabilities.
In each case, \sys serves as a tool for localized, correctness-sensitive transformations with minimal disruption to surrounding logic.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module—where transformation goals can be explicitly stated and verified in isolation.

@@

For instance, a plan for regenerating a specific function might include:
\begin{itemize}
  \item $\mathsf{do}(\texttt{language}(f, \texttt{haskell}))$ — the function must be rewritten in Haskell,
  \item $\mathsf{do}(\texttt{pure}(f))$ — the function must be side-effect-free,
  \item $\mathsf{do}(\texttt{signature}(f))$ — the original function signature must be preserved.
\end{itemize}
% These constraints are scoped to the transformation unit and inform the synthesis phase directly.

\heading{LLM-guided regeneration}
The synthesis task is delegated to a language model guided by a structured prompt. The prompt is constructed by composing fragments contributed by plugins that are part of the regeneration plan, each of which can contribute natural language instructions relevant to its domain. For example:
\begin{itemize}
  \item A signature plugin may contribute: \emph{\say{The output should implement a function \texttt{sqrt(n: float): float}.}}
  \item A length plugin combined with a minimization goal may prepend: \emph{\say{The output should be as short as possible while preserving the original functionality.}}
  \item A language plugin may prepend: \emph{\say{The output function should be written in Haskell.}}
\end{itemize}

@@

% In sum, the synthesis plan acts as a structured contract between logical property reasoning and generative synthesis. By operating at a localized level and under clearly scoped constraints, \sys delivers regenerations that are both controlled and scalable—when applied thoughtfully and systematically.

@@

\heading{Verification as property checking}
After synthesis, \sys invokes a set of property-specific checkers. Each checker is responsible for evaluating whether a property $X$ in $\mathsf{do}(X)$ holds for the regenerated program fragment. These checkers operate at varying levels of precision—ranging from structural pattern checks to light static or symbolic analysis—and are tailored to the property type.

@@

\heading{Targeted regeneration}
\sys is designed as a surgical tool for targeted regeneration tasks, not for whole-program synthesis or full language migration.
It operates most effectively on localized units of code—such as individual functions, methods, or small modules, where behavior is self-contained and transformation goals can be expressed through explicit constraints.
% Typical applications include rewriting high-risk, deprecated, or poorly understood code segments where correctness is critical and semantic preservation is essential.
Broad, untargeted use may yield incomplete results and can be bottlenecked by the limited context of the back-end model \sys uses.

@@

\heading{Language-specific support}
Although some components of \sys are language-agnostic (\eg logic-based planning), others—particularly extraction and verification plugins are not.
Supporting a new language requires implementing language-specific analyzers and constraints. This modularity is deliberate, but it does limit out-of-the-box generality. The current \sys prototype supports Python, JavaScript, Haskell, and C.

>>>>

\begin{abstract}
\end{abstract}
\maketitle

@@

Recent advances in large language models (LLMs) have opened new possibilities for automating program transformation and synthesis tasks, making them attractive tools for software evolution, albeit with caveats.
This flexibility has positioned them as attractive tools for software evolution.
LLMs can translate code across languages~\cite{ou2025enhancingllmbasedcodetranslation},
refactor complex modules~\cite{ziftci2025migrating},
and even generate entire components from scratch~\cite{huynh2025largelanguagemodelscode}.
However, their outputs are shaped by surface-level patterns in the input~\cite{yang2025evaluatinggeneralizationcapabilitieslarge},
they struggle with negative constraints~\cite{hwang2024thinkpinkelephant,jiang2024llmsdreamelephantswhen},
often overfit to spurious input cues~\cite{xu2023llmfoolitselfpromptbased, wu2023deceptpromptexploitingllmdrivencode},
and lack mechanisms for enforcing precise behavioral properties on the generated code~\cite{roh2025breakthechainreasoningfailuresllms}.
As a result, developers must prompt and steer them through brittle heuristics or manual trial-and-error, leading to workflows that are difficult to audit, integrate, or systematically reason about.
Bridging this gap calls for a declarative interface---one that allows developers to specify explicit semantic constraints and transformation goals, and to reason about their satisfaction in a principled way.

@@

@@

\heading{Deployment scenarios and limitations}
\sys can be deployed at various points in the software lifecycle.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating third-party modules that must be adapted to local conventions or hardened against security vulnerabilities.
However, \sys is not designed for whole-program synthesis or large-scale migrations.
Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
To this end, \sys is designed not only as a standalone tool but also as a library component within larger, compound AI systems~\cite{compound-ai-blog}, where it can be invoked as a programmable, constraint-guided regeneration module.

@@

For instance, a plan for regenerating a specific function $f$ might include:
\(\mathsf{do}(\texttt{language}(f, \texttt{haskell}))\) the function must be written in Haskell, 
\(\mathsf{do}(\texttt{pure}(f))\) the function must be side-effect-free, and 
\(\mathsf{do}(\texttt{signature}(f))\) the function signature must be preserved.
% These constraints are scoped to the transformation unit and inform the synthesis phase directly.

\heading{LLM-guided regeneration}
The synthesis task is delegated to a language model guided by a structured prompt. The prompt is constructed by composing fragments contributed by plugins that are part of the regeneration plan, each of which can contribute natural language instructions relevant to its domain. For example:
\begin{itemize}
  \item A \ttt{signature} plugin may contribute: \emph{\say{The output should implement a function \texttt{sqrt(n: float): float}.}}
  \item A \ttt{length} plugin combined with a minimization goal may prepend: \emph{\say{The output should be as short as possible while preserving the original functionality.}}
  \item A \ttt{javascript2haskell} plugin may prepend: \emph{\say{The output function should be written in Haskell.}}
\end{itemize}

@@

% In sum, the synthesis plan acts as a structured contract between logical property reasoning and generative synthesis. By operating at a localized level and under clearly scoped constraints, \sys delivers regenerations that are both controlled and scalable---when applied thoughtfully and systematically.

@@

\heading{Verification as property checking}
After synthesis, \sys invokes a set of property-specific checkers. Each checker is responsible for evaluating whether a property $X$ in $\mathsf{do}(X)$ holds for the regenerated program fragment. These checkers operate at varying levels of precision---ranging from structural pattern checks to light static or symbolic analysis---and are tailored to the property type.

@@

\heading{Targeted regeneration}
\sys is designed as a surgical tool for targeted regeneration tasks, not for whole-program synthesis or full language migration.
It operates most effectively on localized units of code---such as individual functions, methods, or small modules, where behavior is self-contained and transformation goals can be expressed through explicit constraints.
% Typical applications include rewriting high-risk, deprecated, or poorly understood code segments where correctness is critical and semantic preservation is essential.
Broad, untargeted use may yield incomplete results and can be bottlenecked by the limited context of the back-end model \sys uses.

@@

\heading{Language-specific support}
Although some components of \sys are language-agnostic (\eg logic-based planning), others---particularly extraction and verification plugins are not.
Supporting a new language requires implementing language-specific analyzers and constraints. This modularity is deliberate, but it does limit out-of-the-box generality.
Work on a unified language representation~\cite{koppel2018onetool,bap2011,dillig2009sail}, is relevant to overcomming this limitation.
The current \sys prototype supports Python, JavaScript, Haskell, and C.
